---
title: "Change in Types of Crime Before and During the COVID-19 Pandemic"
author: "Kaila An"
date: "`r Sys.Date()`"
output: 
  pdf_document: default
  html_document: FALSE
---

```{r setup, message=FALSE, echo=FALSE, warning=FALSE}
library(data.table)
library(tidyverse)
library(dplyr)
library(plotly)
library(DT)
library(knitr)

# Initialize code chunk options
opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  eval=TRUE,
  echo = TRUE,
  cache = FALSE,
  fig.width = 7, 
  fig.align = 'center',
  fig.asp = 0.618,
  out.width = "700px")
```

## Introduction

  The COVID-19 pandemic had changed the world as we knew it like no other occurrence before. It changed people’s lifestyles, their habits, and overall created a “new normal”. People’s mindsets and priorities have changed so drastically, that it would be interesting to note whether or not it brought a change to criminal behavior and activity. Ever since the start of the pandemic in early 2020, we have seen high rates of crime, like the widespread looting, racial hate crimes, and much more. For this project, I chose to focus on Los Angeles City Data, specifically, ‘Crime Data in LA from 2020 to present’ and ‘Crime Data in LA from 2010 to 2019’. These particular data sets document incidents of crime, transcribed from crime reports. This data has been provided by the Los Angeles Police Department. Based on these data sets, the question I would like to address is:

How has the prevalence of certain types of crimes in Los Angeles changed since the start of the COVID-19 pandemic?

  This final report will examine how the prevalence of certain crimes have changed in LA county as a result of the pandemic and will do so through designing a number of interactive figures that will make it easier to visualize crime instances and characteristics of certain crimes that have taken place. It will also make it possible to detect and analyze certain patterns and contingencies that may be apparent.
  
  Through this report, if we are able to identify certain types of crimes and deduce why they may occur and how new behaviors arising from the start of the pandemic may have contributed to them, it may be possible to assert preventative measures in order to reduce these prevalent crime types.  


## Methods

**How to download data:**
  I obtained the data from the Los Angeles City Data website.  The datasets were fairly large, and so one would need to download them on their own in order to view and manipulate the files.  Below, are steps to downloading the datasets and reading them into R:

1.	Go to the following links and click ‘export’ and then ‘csv’ in order to download the datasets as CSV files directly onto your computer:

  <https://data.lacity.org/Public-Safety/Crime-Data-from-2010-to-2019/63jg-8b9z>

  <https://data.lacity.org/Public-Safety/Crime-Data-from-2020-to-Present/2nrs-mtv8>

2.	Rename the datasets to: “crimedata_2020topresent.csv” and 
“crimedata_2010to2019.csv”
3.	Then, you can read in the data using the codes a few lines below with the data.table::fread function.

**Data cleaning and wrangling:**
  The datasets are extremely large with over a million entries each and 28 variables.  Therefore, I decided to focus on a few in particular when conducting analyses and creating figures.  I focused heavily on 'Crime Code Description', 'Date of Occurrence', and 'Region/Area'.  I delved deeper into these variables and looked into the different categories within them in order to determine changes in prevalence in crime before and after the start of COVID-19.

  I took the 'DATE OCC' or 'date occurred' variable and converted that into year, so that each incidence of crime could be identified and categorized by the year it occurred (anywhere ranging from 2010 to 2022).  I went further as to pick two specific crime types to focus on, Vandalism and Stolen Vehicles.  With these variables, I grouped incidences of these crimes, separately, by year, so that we may be able to observe trends throughout the years.

  I also created tables in which I could find the top 5 crime types for each dataset and considered the 2010-2019 as "pre-COVID-19" and the 2020-present dataset as "post-COVID-19.  After this, I manipulated the data so that I could see how prevalent the top 5 crimes were in different areas of Los Angeles county, pre and post-COVID-19.

**Tools used for data exploration:**
  I utilized packages data.table, tidyverse, dplyr, plotly, DT, stringr, and knitr, in analyzing the data and creating corresponding tables and figures.  


```{r, echo=FALSE, include=FALSE}
# Read in the data
crime201019 <- data.table::fread("crimedata_2010to2019.csv")
crime202022 <- data.table::fread("crimedata_2020topresent.csv")
crime <- rbind(crime201019, crime202022)
```


```{r, echo=FALSE, include=FALSE}
# Check the data
str(crime202022)
str(crime201019)

dim(crime201019)
dim(crime202022)

head(crime201019)
head(crime202022)

tail(crime201019)
tail(crime202022)
```


## Results

**Tables** 
These tables highlight the top 5 crimes for the two different datasets: pre-COVID-19 (2010-2019) and post-COVID-19 (2020-present).  It is important to note that the numbers for table #2 are significantly smaller than that of table #1.  This is due to the fact that table #1 deals with 10 years of crime data and incidences while table #2 only has over 2 years of data to refer to.

```{r, echo=FALSE}
CrimeType <-
  crime201019 %>%
  count(`Crm Cd Desc`)

CrimeType %>%
  arrange(desc(n)) %>%
  top_n(5) %>%
knitr::kable(caption = "Prevalence of Different Types of Crimes in LA County 2010-2019")
```

```{r, echo=FALSE}
CrimeType <-
  crime202022 %>%
  count(`Crm Cd Desc`)

CrimeType %>%
  arrange(desc(n)) %>%
  top_n(5) %>%
knitr::kable(decreasing = TRUE, caption = "Prevalence of Different Types of Crimes in LA County 2020-Present")
```

**Figure 1. Crime Occurrence by Year**

  This interactive bar chart shows the number of crimes (total) that took place yearly from 2010-2022.  We see that rates of crime have remained relatively similar, but there is a spike in crime between 2020 to 2021 (during the COVID-19 pandemic).  There seems to be significantly less crime in 2022.  This is not due to the fact that crime rates had decreased.  It is important to note that the year 2022 is not over yet.  It would be worthwhile to revisit this data set at the end of the year and run the same analysis, in order to see the ending total of crime occurrences for 2022.  Overall, we do not see an all-time high in crime through any of the COVID-19 years, meaning that the pandemic did not bring upon higher rates of crime, in general, which is a positive thing to note.  This means that the most change must have occurred due to other specific characteristics, for which we will examine further in the following figures.

```{r, echo=FALSE, include=FALSE}
library(stringr)
crime$year <- str_extract(crime$`DATE OCC` , "[0-9]{4}") %>%
  as.numeric()
```

```{r, echo=FALSE}
p <- ggplot(crime , aes(y = year, x = ..count.., fill = as.factor(year))) + geom_bar() + guides(fill=guide_legend(title = "Year")) + labs(y = "Year" , x = "# of Occurrence of Crime")
ggplotly(p)
```


**Vandalism (Felony)**

  A new crime type that arose in the Top 5 of 2020-2022 (during the COVID-19 pandemic) was Vandalism.  This would be considered a felony for offenses that were $400 in damages and over and included all church vandalisms.  This is a scatterplot that represents occurrences of these sorts of vandalisms throughout the years.  There is a red line to indicate the start of the COVID-19 pandemic, to make it easier to see where a difference lies, if any.  Vandalism occurrences were on the rise since the start of the pandemic and reached a high in 2021.  The sudden drop in 2022 can also be attributed to the fact that the year 2022 is not over yet.  It's important to note that these Vandalisms could have been attributed to the looting and property damage that was occurring at the time of certain activist movements, such as Black Lives Matter.  The pandemic brought upon many social justice issues, and a lot of people turned to crime as a way to express their anger and to push for their cause.

**Figure 2. Occurrences of Vandalism Before and After COVID-19**
```{r, echo=FALSE}
p <- crime %>% filter(`Crm Cd Desc` == "VANDALISM - FELONY ($400 & OVER, ALL CHURCH VANDALISMS)") %>%
  group_by(year) %>%
  summarise(n = n()) %>%
ggplot(aes(x = year , y = n)) + geom_point() + geom_line() + geom_vline(xintercept = 2020 , color = "red") + scale_x_continuous(breaks = seq(2010, 2022, 1)) + labs(y = "# of Vandalism Occurrences" , x = "Year")
ggplotly(p, width = 500 , height = 200)
```

**Stolen Vehicles**

  The #1 most common crime after the start of the COVID-19 pandemic was stolen vehicles.  This is a scatterplot that represents occurrences of these sorts of vehicle theft throughout the years 2010-2022.  There is a red line to indicate the start of the COVID-19 pandemic, to make it easier to see where a difference lies, if any. We see that vehicle theft was experiencing a steep upwards trend since 2019 and reached an all-time high in 2021.  It seems as though many crime types experienced all-time highs in 2021.  The sudden drop in 2022 can also be attributed to the fact that the year 2022 is not over yet.  Vehicle Theft rising to be the #1 most common crime throughout the pandemic can be attributed to several reasons.  For one, a lot of "hacks" have been circulating on social media regarding how to break into certain models of cars with USB cables.  Also, the pandemic resulted in a lot of layoffs, due to lack of business and owners being unable to pay their employees.  This would often result in businesses shutting down.  In addition, the "work from home" culture brought upon less of a need for certain roles, which led to employees being terminated.  This would result in financial burden, which may have driven those desperate enough to steal vehicles.  

**Figure 3. Occurrences of Stolen Vehicles Before and After COVID-19**
```{r, echo=FALSE}
p <- crime %>% filter(`Crm Cd Desc` == "VEHICLE - STOLEN") %>%
  group_by(year) %>%
  summarise(n = n()) %>%
ggplot(aes(x = year , y = n)) + geom_point() + geom_line() + geom_vline(xintercept = 2020 , color = "red") + scale_x_continuous(breaks = seq(2010, 2022, 1)) + labs(y = "# of Stolen Vehicle Occurrences" , x = "Year")
ggplotly(p, width = 500 , height = 200)
```

**Top 5 Crimes by Area**

  The following figures shows the differing numbers of occurrence of the top 5 crimes for each area of LA county.  There are two tabs that represent the top 5 crimes for the areas for 2010-2019 and the same for 2020-present.  Through this figure, we can see how certain types of crimes may be more prevalent in certain areas than others and if there were overall more or less crime occurrences in different areas of LA county.  One can also navigate whether or not trends of crime significantly changed in these specific areas following the start of the COVID-19 pandemic.  Through these figures, it would be interesting to note how top 5 crimes may have changed before and after the start of the pandemic per region.  One could also deduce whether certain crimes remain high in specific areas, compared to others,  and why this may be.  For example, the regions 'Wilshire' and 'West LA' experience significantly less crime overall than that of 'Central', 'North Hollywood', or 'Hollywood'.  On top of that, the top crime for each of these areas vary greatly.  

**Figure 4. Top 5 Crimes by Area (2010-19)**

```{r echo=FALSE}
p1_scatter <- crime201019 %>% filter(`Crm Cd Desc` %in% c("BATTERY - SIMPLE ASSAULT" , "BURGLARY FROM VEHICLE" , "VEHICLE - STOLEN" , 
"THEFT PLAIN - PETTY ($950 & UNDER)" , "BURGLARY")) %>%
  group_by(`AREA NAME`, `Crm Cd Desc`) %>%
  summarise(n = n()) %>%
ggplot(aes(x = `Crm Cd Desc` , y = n , col = `Crm Cd Desc`)) + geom_point() + facet_wrap(~`AREA NAME` , ncol = 3) + theme(axis.text.x = element_blank()) + labs(col = "Crime Type")
ggplotly(p1_scatter , width = 800 , height = 1000)
```

**Figure 5. Top 5 Crimes by Area 2020-Present**

```{r echo=FALSE}
p2_scatter <- crime202022 %>% filter(`Crm Cd Desc` %in% c("BATTERY - SIMPLE ASSAULT" , "BURGLARY FROM VEHICLE" , "VEHICLE - STOLEN" , 
"VANDALISM - FELONY ($400 & OVER, ALL CHURCH VANDALISMS)" , "BURGLARY")) %>%
  group_by(`AREA NAME`, `Crm Cd Desc`) %>%
  summarise(n = n()) %>%
ggplot(aes(x = `Crm Cd Desc` , y = n , col = `Crm Cd Desc`)) + geom_point() + facet_wrap(~`AREA NAME` , ncol = 3) + theme(axis.text.x = element_blank()) + labs(col = "Crime Type")
ggplotly(p2_scatter , width = 800 , height = 1000)
```

## Conclusion and Summary

  We see that based on the results, rates of crime did spike in the middle of the COVID-19 pandemic, however, there was nothing outstanding compared to the pre-COVID years.  However, we do see that the nature of crime types did, indeed, change following the start of the pandemic.  When we look at the top 5 crime types for both time periods, it is important to note that for the most part, crime types stayed the same.  But, following the start of pandemic, Vandalisms rose to the top 5 list.  Vandalisms were not in the top 5 for the pre-pandemic dataset.  This particular type of vandalism would be considered as a felony and results in $400 and up in damages and includes all church vandalisms.  In addition, we see that the #1 most common crime changed from being battery by simple assault to becoming stolen vehicles after the start of the pandemic.  
  
  It is important to reflect on the reasons why these changes in trends in crime types occurred.  The rise in usage of social media platforms have spread information rapidly like never before.  The app 'Tik Tok' is especially popular in spreading news, trends, and information.  It gained a lot of popularity during the COVID-19 pandemic because everyone was in quarantine, unable to go out as they did before.  There was a recent trend in the past year regarding using USB cables to break into and steal certain car models, such as Hyundai Elantras, which led to a spike in car thefts.  The pandemic led to a lot of unemployment, as well.  If certain struggling people were going through desperate times, they may have turned to theft as a desperate measure, as well.
  
  In addition, Vandalism becoming one of the top 5 most prevalent crimes after the start of the COVID-19 pandemic could have been a result of people being restless and antsy, having to stay at home.  There were also a lot of protests and social justice actions taking place after the start of the pandemic.  This includes, but is not limited to the Black Lives Matter Movement, Roe vs. Wade overturn, etc.  As a result, some people acted radically, turning to vandalism and destroying property in upset.  
  
  When looking towards how the top 5 crime types may have changed and how they affected specific regions of LA county, we see that the top crimes per regions changed greatly following the start of the pandemic.  For example, Central region of LA saw a great increase in burglaries from vehicles after 2020, but in 2010-2019, the top crime was battery by simple assault.  It's also interesting to note that the #1 top crime in almost every region was stolen vehicles.
  
  Overall, the nature and prevalence of types of crime certainly changed after the start of the COVID-19 pandemic.  This could have been due to a plethora of reasons.  For example, the rates of battery by simple assault may have decreased because everyone had been in quarantine for so long during the first half of the pandemic.  There was so little human interaction going on during that period, that it made for batteries by simple assault to be less likely to occur.  In addition, there were social distancing policies put into place, which prevents heavy interaction with strangers.  Further, the rise in rates of different types of burglaries and car theft can be due to the desperation and financial struggle people may have felt more heavily due to lay-offs and rise in unemployment rates that resulted from the pandemic.  The "new normal" brought upon remote work and proved that there was less of a need for certain roles, which led to more layoffs.  Companies, especially small businesses, also struggled to pay their employees due to slower business.  Many businesses even had to permanently or temporarily close down due to lack of customers and revenue.  It could be valuable to run this dataset again, at the end of the year to see the final count of crime for 2022 and see if any trends change or any new ones arise.  We can continue to track crime rates and trends as the pandemic progresses and determine whether crime changes as the effects of the pandemic begin to alleviate.
